Scaling up a GF grammar for free parsing

### 2011-12-09 doc/report/report.tex
report with notes from meeting

Could varandra be thought of as reflexive for 'oss'?
%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

@@

\subsection{Grammatical Framework}
>> extensnion to CFG, same expressiveness as TAG and HG. parse in polynomial time.
>> incremental parsing.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars.

@@

The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
shared by many languages and can be implemented by a numeber of concrete grammars,
each representing one language. The abstract syntax models the semantics of the
domain without concern for language specific features such as agreement or word
order.

@@

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about word forms that are not used.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum %\cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the somewhat incorrect singular form ``glasöga" (``glass-eye").
The same problem occurred for verbs that always are used in passive form:
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

@@

%%PASSIV
\subsubsection{Passive}
>> why not VPSlash here??
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using a modale auxiliary verb \\
\enumsentence{ \shortexnt{6}
{Bilden &blev& tagen& av & rymdsonden & Galileo}
{The picture & was & taken & by & the spacecraft & Galileo}}
\label{gfPass:peri-pass}
or by adding an \emph{s} to
the verb: \\
\enumsentence{\shortex{5}
{Bilden & togs&  av & rymdsonden&  Galileo}
{The picture&  took+\textbf{s}&  by&the  spacecraft& Galileo}
{``The picture was taken by the spacecraft Galileo"}}
\label{gfPass:s-pass}
This second variant, the \textbf{s-passive}, is commonly used in Swedish, some studies
suggest that it is used in more than 80\% of the times \cite{laanemets}.
It is not as common in the other Scandinavian languages however, 
and it can not be used in all tenses for all words. The Norwegian 
translation of both sentence \ref{gfPass:peri-pass} and sentence  \ref{gfPass:s-pass} is
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{Bildet ble tatt av romsonden Galileo}
The resource grammar for Scandinavian therefore expressed the passive using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         ta -> blev tagen| \\
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastic passive is still allowed, but as an alternative rather
than the default.
The grammar also allows using three-place verbs \verb|V3| , as \emph{give}, to
form passives:
\enumsentence{\shortex{6}
{Betyget & 3 &ges &till &många &elever}
{Grade & 3 & give+\textbf{s}& to & many & students}
{``Grade 3 is given to many students"}}
\label{ex:passV3}
by letting the first object be `removed' and given by the subject. 
A remark can be made about sentences where the second object is put first in the sentence, like 
\enumsentence{Till många elever ges betyget 3}
This is a variant of example \ref{ex:passV3}, \emph{`många elever'} is still the object but put in focus. The grammar
parse this by applying \verb|FocObj|.
%%

@@

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').
>> subject controlling verbs
>> why comlicated with doubled types, show simplicity (elegant) with dep.types, and
>> some problems (many fields).
>> samesame as Obj, but conjunction works too, sin is Quant but what i sig?
>> (show circle (sig : NP -> GenNP sig -> sin??)
% agreement to subj: Also good for double functions : Flickan, hon var inte dum, hon?
% F1,2,3 : NP -> VP -> Cl ; NP needs gender (Masc,Fem,Utr,Neutr)

@@

Kanske mindre om detta, skriv om vad jag nu la till.
om 'de äpplen', which are also used in 'finnas'
Since those constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. To fix this, an extra parameter was needed to tell whether
\emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
An more natural sounding alternative would be
\enumsentence{Jag vill ha en sådan katt som inte fäller.}
and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
needed.
du är här, vilket jag vet.
i de äpplen/det äpple du äter bor en mask/ som du ser.\\

>>>>

Mention varandra. Same idea.
%Basic info about Swedish. V2 lang, inverted word order and subordinate clauses.
%Passive, reflexive 'sitt', 'det är', 'det sitter en katt där', prepare the reader
%for what will be written in the grammar part.
%Adjectives but not verbs congruate with nouns. De är stora, They are big, Jag
%är stor, I am big
%Since the s-passive is the most common one in written Swedish \cite{SAG-34-1}, it has

@@

\subsection{Grammatical Framework}
>> extensnion to CFG, same expressiveness as TAG and HG. parse in polynomial time.
>> incremental parsing.
>> how parse trees work, not correct complete representation.
>> not a constituent structure. compare to lfg, fstructure, cstructure.
>> explain how they are created, graphviz and how a user uses them.
The main component of the the project is the Grammatical Framework\cite{gfbok} (GF). It is
a grammar formalism based on Martin-Löf type theory\cite{martinlof}. GF is
designed for multilingual applications and the formalism is equivalent to
Parallel Multiple Context-Free Grammars (Seki \& al. 1991), it is hence
stronger than mildly contex-free grammars.

@@

The key feature in GF is the distinction between
\textit{abstract} and \textit{concrete} syntaxes. The abstract is meant to be
>> internal structure
>> compare fstructure in LFG
shared by many languages and can be implemented by a numeber of concrete grammars,
each representing one language. The abstract syntax models the semantics of the
domain without concern for language specific features such as agreement or word
order.

@@

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about word forms that are not used.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum %\cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the somewhat incorrect singular form ``glasöga" (``glass-eye").
The same problem occurred for verbs that always are used in passive form:
>> not passive!! deponent, s-verbs!!
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

@@

%%PASSIV
\subsubsection{Passive}
>> why not VPSlash here??
There are two ways of forming passive verb phrases in Swedish: the 
\textbf{periphrastic passive}, formed by using a modale auxiliary verb \\
\enumsentence{ \shortexnt{6}
{Bilden &blev& tagen& av & rymdsonden & Galileo}
{The picture & was & taken & by & the spacecraft & Galileo}}
\label{gfPass:peri-pass}
or by adding an \emph{s} to
the verb: \\
\enumsentence{\shortex{5}
{Bilden & togs&  av & rymdsonden&  Galileo}
{The picture&  took+\textbf{s}&  by&the  spacecraft& Galileo}
{``The picture was taken by the spacecraft Galileo"}}
\label{gfPass:s-pass}
This second variant, the \textbf{s-passive}, is commonly used in Swedish, some studies
suggest that it is used in more than 80\% of the times \cite{laanemets}.
It is not as common in the other Scandinavian languages however, 
and it can not be used in all tenses for all words. The Norwegian 
translation of both sentence \ref{gfPass:peri-pass} and sentence  \ref{gfPass:s-pass} is
%The periphrastic passive is preferred, especially in spoken langauge.
\enumsentence{Bildet ble tatt av romsonden Galileo}
The resource grammar for Scandinavian therefore expressed the passive using auxiliary verb. \\
\verb|PassV2 : V2  -> VP | \\
\verb|         ta -> blev tagen| \\
>> frånta, erbjuda works. add. check vp in urdu
>> how are they used with reflexives??
This rules allows two-place verbs to be used in passive by using \emph{bli} (\emph{become}), and thereby
turned into complete verb phrases; they no longer need an object.
The s-passive has now been added as the standard case for the Swedish GF
grammar. Periphrastic passive is still allowed, but as an alternative rather
than the default.
The grammar also allows using three-place verbs \verb|V3| , as \emph{give}, to
form passives:
\enumsentence{\shortex{6}
{Betyget & 3 &ges &till &många &elever}
{Grade & 3 & give+\textbf{s}& to & many & students}
{``Grade 3 is given to many students"}}
\label{ex:passV3}
by letting the first object be `removed' and given by the subject. 
A remark can be made about sentences where the second object is put first in the sentence, like 
\enumsentence{Till många elever ges betyget 3}
This is a variant of example \ref{ex:passV3}, \emph{`många elever'} is still the object but put in focus. The grammar
parse this by applying \verb|FocObj|.
%%

@@

The kind of noun phrases that can be used seems to coincide with the ones
that can be used with formal subjects (see section `Impersonal constructions').
>> subject controlling verbs
>> why comlicated with doubled types, show simplicity (elegant) with dep.types, and
>> some problems (many fields).
>> samesame as Obj, but conjunction works too, sin is Quant but what i sig?
>> (show circle (sig : NP -> GenNP sig -> sin??)
>> how we can add dependent type for Num to use 'varandra'
>> CN should have dettyp too, 'avståndet till sin bil'
% agreement to subj: Also good for double functions : Flickan, hon var inte dum, hon?
% F1,2,3 : NP -> VP -> Cl ; NP needs gender (Masc,Fem,Utr,Neutr)

@@

A few problematic things can be illuminated:
Since the constructions are inspired by other languages, some of them may sound quite
non-natural in Swedish, such as the \verb|RelS|. This combines a sentence \emph{`Hon sover'}
and a relative sentence \emph{`som/vilket är bra'} which wrongly expressed 
\emph{`She sleeps, which is good'} as
\emph{`Hon sover som är bra'}. The error comes from the fact that \emph{`som'} is
embedded in the relative clause. This is easily fixed, by adding an extra
parameter was needed to tell whether \emph{`som'} or \emph{`vilket'} should be used. \\
Another complication stems from \verb|RelCl| which express the english \emph{`such that'}.
The Swedish version \emph{`sådan att'} sounds stelbent, when used outside of logic books.
\enumsentence{Jag vill ha en katt sådan att den inte fäller.} 
This relative clause in the sentence is, in GF, put together by two constituents: \emph{`` katt"}
of type \verb-CN- and \emph{``den inte fäller} of type \verb-RCl- (relative clause). They are
glued together by \emph{``sådan att"} and \emph{``katt"} should agree with \emph{``den"}. 
\enumsentence{*Jag vill ha en katt sådan att \textbf{det} inte fäller.} \label{sent:detfaller}
However, this is not the case and sentence \ref{sent:detfaller} is currently accepted by the grammar.

The grammar has been extended to accept
\enumsentence{\shortex{7}
{\textbf{De uppfattningar}& som& förs& fram ...} 
{The opinions& that& put & forward ...}
{``The opinions presented ..."}}
where \emph{``de"} is used together with a indefinite noun. This is only correct for 
relative clauses or together with the verb \emph{``finnas"}.
In other cases definitev form must be used, cf. sentence \ref{sent:debocker}.
\enumsentence{\textbf{De uppfattningarna} förs fram.\\
``The opinions are presented."}\label{sent:debocker}
% also ok in finnas!!

%An more natural sounding alternative would be
%\enumsentence{Jag vill ha en sådan katt som inte fäller.}
%and hopefully this will be implemented soon, otherwise: why is it hard and how we can say
%'en katt sådan att det regnar'. It is relative but has on empty field, still congruation is
%needed.
%du är här, vilket jag vet.
%i de äpplen/det äpple du äter bor en mask/ som du ser.\\
